---
title: "Exploring NYC Schools Survey Data"
author: "Chenyu Wang"
data: "May 14, 2020"
output: html_notebook
---

In this project, I will analyze data from the New York City school department to understand whether parent, teacher, and student perception of the following factors affect average school SAT scores (an indicator of academic performace):

1. Safety
2. Engagement
3. Comminication
4. Academics

The data, collected in 2011, are publicly available and can be accessed [here](https://data.cityofnewyork.us/Education/2011-NYC-School-Survey/mnz3-dyi8).
The combined data of NYC school can be found [here](https://data.world/dataquest/nyc-schools-data).

I'll start by installing the pakages that I'll need for this analysis and set my working directory:

```{r}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(ggplot2)
```

I'll now import the dataset into R

```{r}
combined <- read.csv("combined.csv")
survey <- read_tsv("masterfile11_gened_final.txt")
survey_d75 <- read_tsv("masterfile11_d75_final.txt")
```

Not all the information are needed for our analysis, therefore, we will only obtain variables that is important for us, for this session, we will keep only High School info.

```{r}
survey_select <- survey %>%
  filter(schooltype == "High School") %>%
  select(dbn:aca_tot_11)
```

Select columns needed for analysis from survey_d75.

```{r}
survey_d75_select <- survey_d75 %>%
  select(dbn:aca_tot_11)
```

Combine survey and survey_d75 together

```{r}
survey_total <- survey_select %>%
  bind_rows(survey_d75_select)
```

Now we want to combine the survey_total with our original combined dataset together, but first, we need to change the "key" name in survey_total from "dbn" to "DBN"

```{r}
survey_total <- survey_total %>%
  rename(DBN = dbn)

# combine two dataframe together by DBN with a left-join
combined_survey <- combined %>%
  left_join(survey_total, by = "DBN")
```


As we are trying to find out the "do student, teacher, and parent perceptions of NYC school quality appear to be related to demographic and academic success metric", we can use a correlation matrix to find out the relationships between pairs of variables.

```{r}
cor_mat <- combined_survey %>%
  select(avg_sat_score, saf_p_11:aca_tot_11) %>%
  cor(use = "pairwise.complete.obs")
```

In order to better filter out the strong relationship, we can convert this matrix to a tibble

```{r}
cor_tib <- cor_mat %>%
  as_tibble(rownames = "variable")
```

We now trying to find correlations of other variables with `avg_sat_score` that are greater than 0.25 or less than -0.25 (strong correlation).

```{r}
strong_cors <- cor_tib %>%
  select(variable, avg_sat_score) %>%
  filter(avg_sat_score > 0.25 | avg_sat_score < -0.25)
```

To visually explore the relationship, I will create scatter plots to demonstrate.

```{r}
# create a function for the scatter plot
create_scatter <- function(x, y) {
  ggplot(data = combined_survey) +
    aes_string(x = x, y = y) +
    geom_point(alpha = 0.3) +
    theme(panel.background = element_rect(fill = "white"))
}

x_var <- strong_cors$variable[2:5]
y_var <- "avg_sat_score"

map2(x_var, y_var, create_scatter)
```


It would be interesting to see whether parents, students, and teachers have similar perceptions about the four school quality metrics they were surveyd about:
Safety and Respect
Communication
Engagement
Academic Expectation

We first reshape the data so we can investigate differences in student, parent and teacher responses to survey questions.

```{r}
combined_survey_gather <- combined_survey %>%
  gather(key = "survey_question", value = score, saf_p_11:aca_tot_11)

# create new variables "response_type" and "questoin"
combined_survey_gather <- combined_survey_gather %>%
  mutate(response_type = str_sub(survey_question, 4, 6)) %>%
  mutate(question = str_sub(survey_question, 1, 3))

# replace "response_type" with parent, teacher, student and total.
combined_survey_gather <- combined_survey_gather %>%
  mutate(response_type = ifelse(response_type == "_p_", "parent", ifelse(response_type == "_t_", "teacher", ifelse(response_type == "_s_", "student", ifelse(response_type == "_to_", "total", "NA")))))
```

Finally, make a boxplot to see if there appear to be differences in how the three groups of responders answered the four questions.

```{r}
combined_survey_gather %>%
  filter(response_type != "total", ) %>%
  ggplot() +
  aes(x = question, y = score, fill = response_type) +
  geom_boxplot()
```

